\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2024-08-12 paper/main.tex
Updated via the Open Science Framework

\label{modeling}\Copy{modeling}{\textcolor{blue}{The \citet{Lewis2005} model is generally used to predict reading times. Recently, \textcite{mertzen} presented simulations for the present design that predicted approximately equally sized effects of syntactic and semantic interference in the reading times of the critical verb. We found that the reading times at the critical verb were confounded by pre-critical effects, which were best explained by \textit{encoding} interference. Therefore, it makes no sense to compare the results from the present SPR experiment to predictions from the Lewis and Vasishth (2005) model of sentence processing as skilled memory \textit{retrieval}. Instead we use the Lewis and Vasishth (2005) model to generate predictions on the microvolt scale for ERP data. For this we adjust the scaling factor $F$ which usually links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R} to link memory activation to brain responses.} The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a \textcolor{blue}{truncated normal prior distribution ($Normal(0, 0.025)$) with a lower bound of 0 and an upper bound of 0.05. This prior reflects the a priori belief that effects will range from 0 to 5 $\mu V$ with higher likelihood for effect sizes below 1 $\mu V$ (see Figure \ref{fig:model_predictions_ERP}). This relatively uninformative prior was chosen because although there exists no previously published ERP data on the Van Dyke (2007) design it reflects the reasonable range of effect sizes in psycholinguistic ERP research.} After sampling the scaling parameter, simulated data was generated from the model while holding all other parameters in the model at fixed values. \textcolor{blue}{We decided to model N400 amplitude differences because Bayes factors provided stronger evidence for the semantic interference effect on the N400 than on the P600 in the present ERP study. To reflect that larger predicted effect sizes correspond to more negative N400 amplitudes, the predicted effects from the model} were multiplied with $-1$.

\textcolor{blue}{For their recent simulations, \textcite{mertzen} created a version of the Lewis and Vasishth (2005) model that assumes that in the \textcite{vandyke07} design, three cues are used at the verb: \{$\pm$ animacy\}, \{$\pm$ grammatical subject\} and \{$\pm$ same clause\}. The \{$\pm$ same clause\} cue matches only with nouns that occurred in the same clause as the verb. ``The addition of the \{$\pm$ same clause\} cue [was] necessary to identify the correct subject'' and as a means for the model to ``to achieve correct retrieval'' in the condition with high syntactic and high semantic interference \citep[][p. 14]{mertzen}. This model predicted approximately equally sized effects of syntactic and semantic interference. Although it is clear that our data does not support equally sized effects of syntactic and semantic interference, we decided to include their version of the model here because it has previously been claimed to be a good model for the present design. We have implemented two alternative versions of the model. Given the evidence against syntactic interference in our data, it is plausible that, at least in this experimental design and language, syntactic interference plays little to no role. We speculate that this could be due to the parser tracking hierarchical structure and that this hierarchical information is used when searching for a noun in memory -- even more so than suggested by \textcite{mertzen}.} \Copy{clause_tracker}{\textcolor{blue}{So, we implemented a model which puts even more emphasis on hierarchical syntactic structure and thereby basically eliminates syntactic interference.} Specifically, the relevant syntactic cues that might be used at the verb are not \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} -- as implemented by \textcite{mertzen} -- but rather a composite cue \{$\pm$ subject-in-same-clause\}. \textcolor{blue}{As already noted by \textcite{mertzen}, structural information like [+ subject of clause$_i$] is not an inherent feature like for example [+ animate], but within in the cue-based retrieval framework, we can easily assume that such a feature is encoded into memory and can consequently be used to match the retrieval cue \{$\pm$ subject of clause$_i$\}. This idea goes back to the roots of the Lewis and Vasishth (2005) model by making the model ``sensitive to whether or not it is parsing an embedded clause'' \citep[][p.388]{Lewis2005} and entails a simple clause tracking mechanism.} The resulting model assumes two retrieval cues: \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. This proposal is \textcolor{blue}{furthermore} in the spirit of the \textcite{dillon2013} and \textcite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval. In the present case, it is possible that syntactically relevant noun is only the one in the same clause as the verb. \textcolor{blue}{For more discussion of the role of structural retrieval cues, see \textcite{franck2020hierarchical} and \textcite{arnett2017subject}.}}\label{clause_tracker} \textcolor{blue}{The here proposed model using two cues predicts no syntactic interference and no interaction, but semantic interference effects. Lastly, we implemented a model with one cue which unambiguously matches the subject. This cue would be the master composite cue \{$\pm$ animate-subject-in-same-clause\}. This model naturally does not predict any interference effects. It can be regarded the null model or no-interference model.}

It is informative to fit the model\textcolor{blue}{s} to the data and then examine the posterior predictive distribution of the \textcolor{blue}{ERPs} for each effect. 
We generated posterior predictive distributions for the effects of interest by fitting the $F$ parameter to the \textcolor{blue}{ERP} data. To avoid overfitting, we fitted $F$ only to the semantic interference effect (and not the syntactic interference effect or interaction). This was done by implementing an Approximate Bayesian Computation approach \parencite{SissonABC}: the model's predictions are generated iteratively, and only those parameter values for the scaling parameter $F$ are retained that predict values for the semantic interference effect that are close to the observed effect size $d$ in the \textcolor{blue}{ERP} data. `Close' is defined as $d \pm \,2\times SE$ ms; for details of this approach as applied to the cue-based retrieval model, see  \textcite{vasishth2020_abc}.

\textcolor{blue}{Out of 200,000 iterations of ABC algorithm, we obtained only 40,748 samples of $F$ for the 3-cues model, 28,718 for the 2-cues model and 6,615 for the 1-cue model. This is not surprising given
the relatively wide prior on $F$. In future work, the posterior distribution, with range [-0.5, -0.1]\,$\mu V$, could be used as a basis for redefining a tighter prior on the scaling parameter $F$. Figure \ref{fig:model_predictions_ERP} shows the  posterior predictive distributions
of the effects of interest in comparison to the prior predictive distribution and our ERP
data. The prior and posterior predictions from the model with three cues show equally large effects of syntactic and semantic interference and an interaction. In contrast, the model using two cues, including a syntactic cue that utilizes syntactic hierarchy, predicts no effect of syntactic interference and no interaction, but a semantic interference effect. The model using one cue predicts no interference effects. The prior and posterior predictions of all models overlap with the effects of interest in the data. }

\Copy{figpriorposteriordata}{
\begin{sidewaysfigure}[htpb]
    \caption{Prior and posterior predicted effect sizes on the microvolt scale, i.e., for negative ERPs, for syntactic and semantic interference and their interaction \textcolor{blue}{from three versions of the Lewis and Vasishth (2005) cue-based retrieval model.}
    %using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. 
    The data with 95 \% credible intervals are shown in red.}\label{fig:model_predictions_ERP}
        \centering
         \includegraphics[width=\textwidth]{images/ERP_priorpred_vs_postpred_vs_data.png}
\end{sidewaysfigure}
}

\textcolor{blue}{To evaluate whether our new 2-cues model quantitatively outperforms the two other models in its predictive fit, we computed Bayes factors for the relevant model pairs under six different priors for the $F$ parameter 
($Normal$(0.01,0.01), 
$Normal$(0.01,0.02),
$Normal$(0.01,0.005),
$Normal$(0.02,0.01),
$Normal$(0.02,0.02),
$Normal$(0.02,0.005)). First, we computed the likelihood of the sampled $F$ parameters under each of the priors (LKL$_{F-prior}$) and the sampling prior (LKL$_{F-model}$). Then, the likelihood of the predicted semantic interference effect from the model was computed given the size of the observed semantic interference effect (LKL$_{predicted-effect}$). To obtain the marginal likelihood of the model, the product of LKL$_{predicted-effect}$ and LKL$_{F-prior}$ was divided by LKL$_{F-model}$. The marginal likelihood of the 2-cues model was then divided by the marginal likelihood of the three-cues and one-cue model, respectively. The resulting Bayes factors are shown in Figure \ref{fig:BF_compmods}. Compared to the null model with one cue, the Bayes factors provide extreme evidence for the new two-cues model. Therefore, we can conclude that the computational modeling shows that our data shows interference. Compared to the three-cues model proposed by \textcite{mertzen}, the Bayes factors provide moderate evidence for the new two-cues model. Therefore, the assumption that the relevant syntactic cue is the \{$\pm$subject-in-same-clause\} cue – rather than the previously assumed \{$\pm$grammatical subject\} cue – is moderately supported by these modeling results.
}

>>>>

\label{modeling}\Copy{modeling}{\textcolor{blue}{The \citet{Lewis2005} model is generally used to predict reading times. Recently, \textcite{mertzen} presented simulations for the present design that predicted approximately equally sized effects of syntactic and semantic interference in the reading times of the critical verb. We found that the reading times at the critical verb were confounded by pre-critical effects, which were best explained by \textit{encoding} interference. Therefore, it makes no sense to compare the results from the present SPR experiment to predictions from the Lewis and Vasishth (2005) model of sentence processing as skilled memory \textit{retrieval}. Instead we use the Lewis and Vasishth (2005) model to generate predictions on the microvolt scale for ERP data. For this we adjust the scaling factor $F$ which usually links the activation of items in memory to retrieval times \citep{Lewis2005,anderson_etal_2004_ACT_R} to link memory activation to brain responses.} The modeling reported here is, to the best of our knowledge, the first time that the \citet{Lewis2005} model has been fit quantitatively to ERP data. We sampled the free scaling parameter $F$ from a \textcolor{blue}{truncated normal prior distribution ($Normal_{lb=0,ub=0.05}(0.01, 0.01)$) with a lower bound of 0 and an upper bound of 0.05. This prior reflects the a priori belief that effect on the microvolt scale will range from 0 to 5 $\mu V$ with higher likelihood for effect sizes below 3 $\mu V$ (see Figure \ref{fig:model_predictions_ERP}). This relatively uninformative prior was chosen because although there exists no previously published ERP data on the Van Dyke (2007) design it reflects the reasonable range of effect sizes in psycholinguistic ERP research. In future work, the posterior distribution of the semantic interference effect in our ERP data, with range [-0.5, -0.1]\,$\mu V$, could be used as a basis for redefining a tighter prior on the scaling parameter $F$.} After sampling the scaling parameter, simulated data was generated from the model while holding all other parameters in the model at fixed values. \textcolor{blue}{We decided to model N400 amplitude differences because Bayes factors provided stronger evidence for the semantic interference effect on the N400 than on the P600 in the present ERP study. To reflect that larger predicted effect sizes correspond to more negative N400 amplitudes, the predicted effects from the model} were multiplied with $-1$.

\textcolor{blue}{For their recent simulations, \textcite{mertzen} created a version of the Lewis and Vasishth (2005) model that assumes that in the \textcite{vandyke07} design, three cues are used at the verb: \{$\pm$ animacy\}, \{$\pm$ grammatical subject\} and \{$\pm$ same clause\}. The \{$\pm$ same clause\} cue matches only with nouns that occurred in the same clause as the verb. ``The addition of the \{$\pm$ same clause\} cue [was] necessary to identify the correct subject'' and as a means for the model ``to achieve correct retrieval'' in the condition with high syntactic and high semantic interference \citep[][p. 14]{mertzen}. This model predicted approximately equally sized effects of syntactic and semantic interference. Although it is clear that our data does not support equally sized effects of syntactic and semantic interference, we decided to include their version of the model here because it has previously been claimed to be a good model for the present design. We have implemented two alternative versions of the model. Given the evidence against syntactic interference in our data, it is plausible that, at least in this experimental design and language, syntactic interference plays little to no role. We speculate that this could be due to the parser tracking hierarchical structure and that this hierarchical information is used when searching for a noun in memory -- even more so than suggested by \textcite{mertzen}.} \Copy{clause_tracker}{\textcolor{blue}{So, we implemented a model which puts even more emphasis on hierarchical syntactic structure and thereby basically eliminates syntactic interference.} Specifically, the relevant syntactic cues that might be used at the verb are not \{$\pm$ grammatical subject\} and \{$\pm$ same clause\} -- as implemented by \textcite{mertzen} -- but rather a composite cue \{$\pm$ subject-in-same-clause\}. \textcolor{blue}{As already noted by \textcite{mertzen}, structural information like [+ subject of clause$_i$] is not an inherent feature like for example [+ animate], but within in the cue-based retrieval framework, we can easily assume that such a feature is encoded into memory and can consequently be used to match the retrieval cue \{$\pm$ subject of clause$_i$\}. This idea goes back to the roots of the Lewis and Vasishth (2005) model by making the model ``sensitive to whether or not it is parsing an embedded clause'' \citep[][p.388]{Lewis2005} and entails a simple clause tracking mechanism.} The resulting model assumes two retrieval cues: \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. This proposal is \textcolor{blue}{furthermore} in the spirit of the \textcite{dillon2013} and \textcite{Sturt2003} claim, that the parser intelligently targets only the syntactically relevant noun during antecedent retrieval. In the present case, it is possible that syntactically relevant noun is only the one in the same clause as the verb. \textcolor{blue}{For more discussion of the role of structural retrieval cues, see \textcite{franck2020hierarchical} and \textcite{arnett2017subject}.}}\label{clause_tracker} \textcolor{blue}{The here proposed model using two cues predicts no syntactic interference and no interaction, but semantic interference effects. Lastly, we implemented a model with one cue which unambiguously matches the subject. This cue would be the master composite cue \{$\pm$ animate-subject-in-same-clause\}. This model naturally does not predict any interference effects. It can be regarded the null model or no-interference model.}

\textcolor{blue}{Figure \ref{fig:model_predictions_ERP} shows the  prior predictive distributions of the effects of interest for the three competing models discussed above in comparison to our ERP data.  The null model model using one cue predicts no interference effects. The new two-cues model, including a syntactic cue that utilizes syntactic hierarchy, predicts no effect of syntactic interference and no interaction, but a semantic interference effect. The prior predictions from the three-cues model show equally large effects of syntactic and semantic interference and an interaction. The predictions of all models overlap with the effects of interest in the data. }

\Copy{figpriorposteriordata}{
\begin{figure}[htpb]
    \caption{The prior predictions on the microvolt scale for syntactic and semantic interference and their interaction \textcolor{blue}{from three versions of the Lewis and Vasishth (2005) cue-based retrieval model.}
    %using the two cues \{$\pm$ subject-in-same-clause\} and \{$\pm$ animate\}. 
    The data with 95 \% credible intervals are shown in red.}\label{fig:model_predictions_ERP}
        \centering
         \includegraphics[width=\textwidth]{images/compmod_predictions.png}
\end{figure}
}

\textcolor{blue}{To quantitatively evaluate the relative predictive performances of the models on the semantic interference data, we used a Bayes factor analysis. We computed Bayes factors for the two-cues model compared to the three-cues model and to the one cue model, respectively. Since Bayes factors are sensitive to the choice of priors on the parameters of interest, we computed Bayes factors under six different priors for the latency factor parameter $F$,
$Normal$(0.01,0.01), 
$Normal$(0.01,0.02),
$Normal$(0.01,0.005),
$Normal$(0.02,0.01),
$Normal$(0.02,0.02), and 
$Normal$(0.02,0.005). }

\textcolor{blue}{The procedure for computing Bayes factors is described as follows. The Bayes factor computation required us to estimate the marginal likelihood of the competing models. We used the Monte Carlo integration method to estimate the marginal likelihood of each model. First, we chose an importance density for the parameter $F$ from which we draw $0.2$ million samples of $F$. Then, for each of these proposed parameter values, we compute the likelihood and prior density. The estimated marginal likelihood of a model will be given by}

\begin{equation}
ML = \frac{1}{n} \sum_{i=1}{n} \mathcal{L}{F_i|y} p(F_i)  \text{ where } F_i \sim Normal_{lb=0,ub=0.05}(0.0.025) 
\end{equation}

\noindent \textcolor{blue}{where $y$ is the observed data, i.e., the observed semantic interference effect in our case, $F_i$ is the $i^{th}$ proposal from the importance density $Normal_{lb=0,ub=0.05}(0.0.025) $. The term $\mathcal{L}{F_i|y}$ gives the likelihood of proposal $F_i$ given the data $y$ and $p(F_i)$ gives the prior density of $F_i$. We compute marginal likelihoods under different prior assumptions $p(F)$.}

\textcolor{blue}{The Lewis and Vasishth (2005) model is a complex, non-deterministic model and it is difficult to express the model's likelihood analytically without drastically simplifying the model \parencite[e.g., in][]{NicenboimRetrieval2018,lisson2020computational}. In absence of an analytically-expressed likelihood function, we used an Approximate Bayesian Computation (ABC)~\citep{sisson2018handbook,palestro2018likelihood} approach to estimate approximate likelihood for a given value of $F$. Under this approach, we compared the model generated data conditioned on $F=F_i$ and the actual observed data $y$. If the model-generated data $x_{i} \sim Model(F_i)$ for $F_i$ was close to the actual data $y$, then $F_i$ has higher likelihood and vice versa. The likelihood term can be rewritten as $\mathcal{L}{F_i|y} = \Psi(d(x_i, y), 0 , \delta)$, where $\Psi(.|\delta)$ represent a density kernel that weighs a proposal $F_i$ based on distance between model-generated data and actual data, $d(x_i,y)$.}

%\textcolor{blue}{Using the above method, we estimate marginal likelihood for each candidate model under six different prior assumptions. The Bayes factor in favor of a model M$_1$ compared to a model M$_2$, i.e., BF$_{12}$, is computed as the ratio of the marginal likelihoods of M$_1$ and M$_2$. The Bayes factor BF$_{12}$ represents the extent to which the model M$_1$ is more likely than M$_2$ given the data. Following the convention from \textcite{jeffreys1998theory}, a Bayes factor above 10 is interpreted as strong evidence in favor of M$_1$ and a value between 3 and 10 is interpreted as moderate evidence in favor of M$_1$ (see also Table \ref{tab:bf_interpretation}).}

\textcolor{blue}{From the marginal likelihoods, we computed Bayes factors for the two-cues model compared to the three-cues model and the one-cue model. That is, the marginal likelihood of the two-cues model $\mathrm{M_1}$ was divided by the marginal likelihood of the three-cues model $\mathrm{M_2}$ and one-cue model $\mathrm{M_3}$ to obtain the Bayes factors $BF_{12}$ and $BF_{13}$, respectively. Following the convention from \textcite{jeffreys1998theory}, a Bayes factor above 10 is interpreted as strong evidence in favor of the two-cues model and a value between 3 and 10 is interpreted as moderate evidence in favor of the two-cues model compared to the competing model (see also Table \ref{tab:bf_interpretation}).}

\textcolor{blue}{The Bayes factors under different priors on $F$ are shown in Figure \ref{fig:BF_compmods}. Compared to the null model, the one-cue model, the Bayes factors indicate strong evidence for the new two-cues model. Therefore, we can conclude that the computational modeling shows that our data shows interference. Compared to the three-cues model proposed by \textcite{mertzen}, the Bayes factors provide moderate evidence for the new two-cues model. Therefore, the assumption that the relevant syntactic cue is the \{$\pm$subject-in-same-clause\} cue – rather than the previously assumed \{$\pm$grammatical subject\} cue – is moderately supported by these modeling results. This result holds for prior assumptions that allows higher probability mass for small values of $F$., i.e., for $F<0.01$.}
